\textls[-0]{\thetitle}

### 2023-03-12 metatheory-condensed.tex
Partially Changing to named variables. For FMLTT. Please review.

% \input{mltt-selected.tex}
\input{mltt-selected-named.tex}
\input{fmltt-selected.tex}

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of (predicative) MLTT
(and \cref{fig:fmltt-selected} \TT). We formulate in an intrinsic style---every appearing syntax piece
is well-typed in prior, and the premise judgement required for well-typed might be omitted for conciseness. 
For example, the well-typed context $\goodCtx{\Gm}{}$ required by \ruleref{n/tm/lam} is omitted.  
We also use round bracket to indicate substitution. 
For each derivation (of each judgement) we introduce judgemental equality indicating convertibility 
in MLTT.  
We have dependent function and dependent tuple type equipped by standard dependent type theory.
%We briefly review the base MLTT fragment first.
%
% In keeping with the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}\YZ{Does coquand2018canonicity really use explicit subst?}\EDJreply{Yes it is. Look at Section 2.1. When he starts reasoning, he actually deals with category with family (instead of the typing rules), where he consider morphism between contexts, which is substitution. But I agree it can be a bit ambiguous to see (because he literally show two version of MLTT without connecting them---one classical typing rule version, the other directly works on Category with Family (CwF)) but since his proof is mainly on Category with Family, so it counts as explicit substitution. But if you are concerned, we can change to Andras Kovacs "Generalized Universe Hierarchies and First-Class Universe Levels". or Andreas Abel's "Decidability of Conversion for Type Theory in Type Theory". The former one directly work on CwF, the latter one include substitution as part of the raw syntax.}
% in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
% substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
% of the syntax rather than meta-operations.
%
% Variables are in the form of \citeauthor{debruijn1964} indices:
% \var{n} is the variable bound by the $n$-th closest enclosing binder.
% For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
% Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
% $\goodSub{\Gm}{\gm}{\Dl}$.
% The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
% yields terms valid in~$\Gm$ (\ruleref{n/tm/sub} and \ruleref{n/ty/sub}).
% %
% The two main forms of substitutions are weakening (\ruleref{n/sub/wk}) and extension (\ruleref{n/sub/ext}):
% \sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
% \sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
% then applies~$\gm$.
% For example, rule \ruleref{n/tm/snd} states that if $t$ is a dependent pair
% that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
% $\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
% where $\SubstId$ is the identity substitution (\ruleref{n/sub/id}).

@@

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{n/tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

@@

We use \cref{fig:stlc-linkage-typing} to illustrate.
On the left is code excerpted from \cref{fig:stlc-mechanized}, and on the right
is how the corresponding fields are modeled and typed in \TT.
%
$\lkg_i$ is the linkage that adds the $i$-th field~$t_i$ with the typing
$\goodTerm{%
\tikzmarkin[disable rounded corners=true,set fill color=A-color,set border color=A-color]{A-i}(-0.02,-0.10)(-0.02,0.28)
A_i%
\tikzmarkend{A-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=t-color,set border color=t-color]{t-i}(-0.02,-0.10)(-0.02,0.28)%
t_i%
\tikzmarkend{t-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=T-color,set border color=T-color]{T-i}(-0.02,-0.10)(-0.02,0.28)%
T_i%
\tikzmarkend{T-i}%
}$.
%
For readability, we use a notation with explicit names
in this typing judgment:
we name the context as $\mathit{self}$ (cf.~\cref{fig:stlc-compiled})
and write its type~$A_i$, which is a dependent tuple type, as a
dependent record type with labels.
%
\begin{itemize}
% [leftmargin=4ex,itemsep=.5ex]
  [labelsep=*,leftmargin=1pc,itemsep=3pt]

@@

The third premise
$\goodTerm{\Gm,\TyTkg{\lsig}}{\seal}{\sub{A}{\SubstWeak{1}}}$
in rules \ruleref{l/add} and \ruleref{lsig/add} is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{tyeq/pk/add})
that packages the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{tmeq/pk/add}) into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.
%
It is straightforward to find the~$\seal$ that fits the bill, though this process
is not automated in \TT.
In particular, when no hiding is needed, $\seal$ is $\var{0}$
%\EDJ{$\var{0}$ is not identity (substitution), but the first variable, right? Or you meant to say it is identity ``sealing''?}\YZreply{Yes, I meant identity sealing. I cannot say 'sealing' because the paper currently does not use the word sealing anywhere.}\EDJreply{What about just remove the parenthesis and ``the identity''}
and $A$ is $\TyTkg{\lsig}$.
In \cref{fig:stlc-linkage-typing}, $\seal_1$, $\seal_2$, and $\seal_{10}$ are $\var{0}$,
and $\seal_6$ is from
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \TyS{\wcode{\wsig_{\texttt{tm}}}},\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$
\endgroup
to
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \cU,\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$%
\endgroup
.

>>>>

% \input{mltt-selected.tex}
\input{mltt-selected-named.tex}
\input{fmltt-selected-named.tex}

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of (predicative) MLTT
(and \cref{fig:fmltt-selected} \TT). We formulate in an intrinsic style---every appearing syntax piece
is well-typed in prior, and the premise judgement required for well-typed might be omitted for conciseness. 
For example, the well-typed context $\goodCtx{\Gm}{}$ required by \ruleref{n/tm/lam} is omitted.  
We also use square bracket to indicate substitution. 
For each derivation (of each judgement) we introduce judgemental equality indicating convertibility 
in MLTT.  
We have dependent function and dependent tuple type equipped by standard dependent type theory.
%We briefly review the base MLTT fragment first.
%
% In keeping with the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}\YZ{Does coquand2018canonicity really use explicit subst?}\EDJreply{Yes it is. Look at Section 2.1. When he starts reasoning, he actually deals with category with family (instead of the typing rules), where he consider morphism between contexts, which is substitution. But I agree it can be a bit ambiguous to see (because he literally show two version of MLTT without connecting them---one classical typing rule version, the other directly works on Category with Family (CwF)) but since his proof is mainly on Category with Family, so it counts as explicit substitution. But if you are concerned, we can change to Andras Kovacs "Generalized Universe Hierarchies and First-Class Universe Levels". or Andreas Abel's "Decidability of Conversion for Type Theory in Type Theory". The former one directly work on CwF, the latter one include substitution as part of the raw syntax.}
% in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
% substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
% of the syntax rather than meta-operations.
%
% Variables are in the form of \citeauthor{debruijn1964} indices:
% \var{n} is the variable bound by the $n$-th closest enclosing binder.
% For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
% Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
% $\goodSub{\Gm}{\gm}{\Dl}$.
% The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
% yields terms valid in~$\Gm$ (\ruleref{n/tm/sub} and \ruleref{n/ty/sub}).
% %
% The two main forms of substitutions are weakening (\ruleref{n/sub/wk}) and extension (\ruleref{n/sub/ext}):
% \sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
% \sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
% then applies~$\gm$.
% For example, rule \ruleref{n/tm/snd} states that if $t$ is a dependent pair
% that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
% $\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
% where $\SubstId$ is the identity substitution (\ruleref{n/sub/id}).

@@

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, a : A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, a : A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{n/tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

@@

We use \cref{fig:stlc-linkage-typing} to illustrate.
On the left is code excerpted from \cref{fig:stlc-mechanized}, and on the right
is how the corresponding fields are modeled and typed in \TT.
%
$\lkg_i$ is the linkage that adds the $i$-th field~$t_i$ with the typing
$\goodTerm{%
\tikzmarkin[disable rounded corners=true,set fill color=A-color,set border color=A-color]{A-i}(-0.02,-0.10)(-0.02,0.28)
a_i : A_i%
\tikzmarkend{A-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=t-color,set border color=t-color]{t-i}(-0.02,-0.10)(-0.02,0.28)%
t_i%
\tikzmarkend{t-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=T-color,set border color=T-color]{T-i}(-0.02,-0.10)(-0.02,0.28)%
T_i%
\tikzmarkend{T-i}%
}$.
%
For readability, we use a notation with explicit names
in this typing judgment:
we name the context as $\mathit{self}$ (cf.~\cref{fig:stlc-compiled})
and write its type~$A_i$, which is a dependent tuple type, as a
dependent record type with labels.
%
\begin{itemize}
% [leftmargin=4ex,itemsep=.5ex]
  [labelsep=*,leftmargin=1pc,itemsep=3pt]

@@

The third premise
$\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$
in rules \ruleref{l/add} and \ruleref{lsig/add} is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{tyeq/pk/add})
that packages the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{tmeq/pk/add}) into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.
%
It is straightforward to find the~$\seal$ that fits the bill, though this process
is not automated in \TT.
In particular, when no hiding is needed, $\seal$ is $\var{0}$
%\EDJ{$\var{0}$ is not identity (substitution), but the first variable, right? Or you meant to say it is identity ``sealing''?}\YZreply{Yes, I meant identity sealing. I cannot say 'sealing' because the paper currently does not use the word sealing anywhere.}\EDJreply{What about just remove the parenthesis and ``the identity''}
and $A$ is $\TyTkg{\lsig}$.
In \cref{fig:stlc-linkage-typing}, $\seal_1$, $\seal_2$, and $\seal_{10}$ are $\var{0}$,
and $\seal_6$ is from
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \TyS{\wcode{\wsig_{\texttt{tm}}}},\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$
\endgroup
to
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \cU,\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$%
\endgroup
.
